# Choosing a generalized linear model with R

---
title: "GLM Analysis of Grass Species Data"
format: html
code-fold: show
---

This document works through building a generalized linear model (GLM) in R using experimental plant ecology data examining grass species response to soil conditions.

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(DHARMa)
library(MASS)

# Set ggplot theme
theme_set(theme_classic())
```

## Data 

We'll analyze an experimental plant ecology dataset examining grass species response to soil conditions with 3 variables and 90 observations:

- **pH** (categorical): Soil acidity level (low, mid, high) - experimental treatment
- **Biomass** (numeric): Plant biomass measurement (likely dry weight in grams)
- **Species** (integer): Number of grass species present in experimental plot

## Step 1: Examine Data

Let's read in the data and examine its structure:

```{r data-load}
dat <- read_csv("Data/grasses.csv")
```

### Data Structure and Summary

```{r data-summary}
# Basic structure
str(dat)

# Dimensions
cat("Dataset dimensions:", nrow(dat), "rows,", ncol(dat), "columns\n\n")

# Summary of each variable
for(var in names(dat)) {
  cat("Variable:", var, "\n")
  cat("Class:", class(dat[[var]]), "\n")
  
  if(is.numeric(dat[[var]])) {
    cat("Type: Numeric values\n")
    cat("Range:", min(dat[[var]], na.rm = TRUE), "to", max(dat[[var]], na.rm = TRUE), "\n")
    cat("Quantiles:\n")
    print(quantile(dat[[var]], na.rm = TRUE))
    
    # Determine numeric type
    if(all(dat[[var]] >= 0, na.rm = TRUE)) {
      if(all(dat[[var]] %% 1 == 0, na.rm = TRUE)) {
        cat("Data type: Positive integers (>= 0)\n")
      } else {
        cat("Data type: Continuous values >= 0\n")
      }
    } else {
      cat("Data type: Continuous values (positive and negative)\n")
    }
  } else {
    cat("Type: Categorical\n")
    freq_table <- table(dat[[var]], useNA = "always")
    cat("Frequency table:\n")
    print(freq_table)
    
    # Determine categorical type
    n_levels <- length(unique(dat[[var]][!is.na(dat[[var]])]))
    if(n_levels == 2) {
      cat("Data type: Categorical, two factors\n")
    } else {
      cat("Data type: Categorical, multiple factors (", n_levels, "levels)\n")
    }
  }
  cat("\n")
}
```

### Data Visualization

```{r data-viz, fig.width=10, fig.height=6}
# Create visualizations for each variable
p1 <- ggplot(dat, aes(x = pH)) +
  geom_bar(fill = "steelblue", alpha = 0.7) +
  labs(title = "Distribution of pH levels", x = "pH Level", y = "Count")

p2 <- ggplot(dat, aes(x = Biomass)) +
  geom_histogram(bins = 20, fill = "forestgreen", alpha = 0.7) +
  labs(title = "Distribution of Biomass", x = "Biomass", y = "Count")

p3 <- ggplot(dat, aes(x = Species)) +
  geom_histogram(bins = max(dat$Species) - min(dat$Species) + 1, 
                 fill = "orange", alpha = 0.7) +
  labs(title = "Distribution of Species Count", x = "Number of Species", y = "Count")

# Relationship plots
p4 <- ggplot(dat, aes(x = pH, y = Species)) +
  geom_boxplot(fill = "lightblue", alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.6) +
  labs(title = "Species count by pH level", x = "pH Level", y = "Number of Species")

library(patchwork)
(p1 | p2) / (p3 | p4)
```

## CHECKPOINT 1 ✓

**Data Interpretation Confirmed:**

1. **pH**: Categorical variable with multiple factors (low, mid, high) - experimental treatment
2. **Biomass**: Continuous numeric values (>= 0, representing dry weight in grams)  
3. **Species**: Positive integers >= 0 (count data representing number of grass species)

---

## Step 2: Build Model Formula

Now we need to determine the structure of our GLM by identifying the response and predictor variables.

### Response Variable Selection

Looking at our experimental design, we have an ecology experiment examining grass species response to soil conditions. The logical response variable would be **Species** (the count of grass species present), as this represents the biological outcome we're trying to understand.

### Predictor Variable Selection

Available predictor variables:
- **pH**: Experimental treatment (soil acidity level)
- **Biomass**: Plant biomass measurement

### Theoretical Justification for Predictors

```{r model-reasoning}
# Let's examine the relationships between variables
cat("Correlation between Biomass and Species:\n")
cor.test(dat$Biomass, dat$Species)

# Examine means by pH level
cat("\nSpecies count by pH level:\n")
dat %>% 
  group_by(pH) %>% 
  summarise(
    n = n(),
    mean_species = mean(Species),
    sd_species = sd(Species),
    mean_biomass = mean(Biomass),
    sd_biomass = sd(Biomass)
  ) %>%
  knitr::kable(digits = 2)
```

### Additional Exploratory Plots

```{r relationship-plots, fig.width=10, fig.height=6}
# Explore relationships between all variables
p1 <- ggplot(dat, aes(x = Biomass, y = Species)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(title = "Species vs Biomass", x = "Biomass", y = "Species Count")

p2 <- ggplot(dat, aes(x = Biomass, y = Species, color = pH)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Species vs Biomass by pH", x = "Biomass", y = "Species Count")

p3 <- ggplot(dat, aes(x = pH, y = Biomass)) +
  geom_boxplot(fill = "lightgreen", alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.6) +
  labs(title = "Biomass by pH level", x = "pH Level", y = "Biomass")

library(patchwork)
p1 / (p2 | p3)
```

## CHECKPOINT 2 ✓

**Model Formula Confirmed:**

**Response variable:** `Species` (count of grass species)

**Model formula:** `Species ~ pH * Biomass`

This interaction model will test:
- Main effect of pH on species count
- Main effect of Biomass on species count  
- Whether the effect of Biomass differs across pH levels

---

## Step 3: Choose Model Family

Now we need to select the appropriate probability distribution (family) for our GLM based on the nature of our response variable.

### Response Variable Analysis

Our response variable `Species` is:
- **Count data**: Positive integers (1, 2, 3, ..., 9)
- **Discrete**: Cannot take fractional values
- **Lower bounded**: Cannot be negative (≥ 0, though our data shows ≥ 1)
- **No obvious upper limit**: Could theoretically be any positive integer

### Model Family Selection

```{r family-analysis}
# Examine the distribution properties of Species
cat("Species count distribution:\n")
table(dat$Species)

cat("\nBasic statistics:\n")
cat("Mean:", mean(dat$Species), "\n")
cat("Variance:", var(dat$Species), "\n")
cat("Variance/Mean ratio:", var(dat$Species)/mean(dat$Species), "\n")

# Check for overdispersion
variance_mean_ratio <- var(dat$Species)/mean(dat$Species)
if(variance_mean_ratio > 1.5) {
  cat("Variance > Mean: Suggests overdispersion (negative binomial may be appropriate)\n")
} else if(variance_mean_ratio < 0.8) {
  cat("Variance < Mean: Suggests underdispersion\n") 
} else {
  cat("Variance ≈ Mean: Poisson distribution may be appropriate\n")
}

# Visualize distribution
hist(dat$Species, breaks = seq(0.5, max(dat$Species) + 0.5, 1), 
     main = "Distribution of Species Count", 
     xlab = "Number of Species", 
     col = "lightblue", 
     border = "black")
```

### Family Decision Reasoning

Based on our analysis:

1. **Data type**: Count data (positive integers) → rules out Gaussian and Gamma families

2. **Distribution characteristics**: 
   - If Variance ≈ Mean → **Poisson** family appropriate
   - If Variance > Mean (overdispersion) → **Negative Binomial** family appropriate
   - If Variance < Mean (underdispersion) → May need zero-inflated or other specialized models

3. **Default choice for count data**: Start with **Negative Binomial** family as it:
   - Handles overdispersion naturally
   - Reduces to Poisson when dispersion parameter = 0
   - More robust for ecological count data

## CHECKPOINT 3 ✓

**Model Family Confirmed:** **Negative Binomial**

**Final Model Specification:**
```r
model <- glm.nb(Species ~ pH * Biomass, data = dat)
```

---

## Step 4: Fit Model and Check Diagnostics

Now we'll fit our negative binomial GLM and assess model assumptions using diagnostic plots.

### Model Fitting

```{r model-fitting}
# Fit the negative binomial GLM
model_nb <- glm.nb(Species ~ pH * Biomass, data = dat)

# Model summary
summary(model_nb)
```

### Model Interpretation

```{r model-interpretation}
# Extract key model information
cat("Model coefficients:\n")
coef(model_nb)

cat("\nTheta (dispersion parameter):", model_nb$theta, "\n")
cat("AIC:", AIC(model_nb), "\n")

# Check significance of interaction
cat("\nANOVA table:\n")
anova(model_nb, test = "Chisq")
```

### Diagnostic Plots with DHARMa

DHARMa (Diagnostics for HierArchical Regression via Monte-carlo) provides residual diagnostics for GLMs through simulation.

```{r dharma-diagnostics, fig.width=12, fig.height=8}
# Create DHARMa residuals
residuals_sim <- simulateResiduals(fittedModel = model_nb, plot = TRUE)
```

```{r additional-dharma-tests}
# Test for dispersion
cat("Dispersion test:\n")
testDispersion(residuals_sim)

cat("\nUniformity test (Kolmogorov-Smirnov):\n")
testUniformity(residuals_sim)

cat("\nOutlier test:\n")
testOutliers(residuals_sim)
```

### Additional Diagnostic Checks

```{r additional-diagnostics, fig.width=12, fig.height=6}
# Residuals vs fitted
par(mfrow = c(1, 2))

# Standard residual plots
plot(fitted(model_nb), residuals(model_nb, type = "pearson"),
     xlab = "Fitted Values", ylab = "Pearson Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

# Q-Q plot of residuals
qqnorm(residuals(model_nb, type = "pearson"))
qqline(residuals(model_nb, type = "pearson"), col = "red")

par(mfrow = c(1, 1))
```

### Model Comparison (Optional)

```{r model-comparison}
# Compare with Poisson model to check if NB was necessary
model_poisson <- glm(Species ~ pH * Biomass, data = dat, family = poisson)

# Compare AIC
cat("Model comparison:\n")
cat("Negative Binomial AIC:", AIC(model_nb), "\n")
cat("Poisson AIC:", AIC(model_poisson), "\n")
cat("Difference (NB - Poisson):", AIC(model_nb) - AIC(model_poisson), "\n")

# Likelihood ratio test
lrtest_result <- anova(model_poisson, model_nb, test = "Chisq")
print(lrtest_result)
```

### DHARMa Package Information

```{r dharma-info}
# Display DHARMa vignette information for user reference
cat("For more information on interpreting DHARMa diagnostics, run:\n")
cat("vignette('DHARMa')\n\n")

cat("Key DHARMa diagnostic interpretations:\n")
cat("- QQ plot should show points along diagonal line\n")
cat("- Residual vs Predicted should show random scatter around horizontal line\n")
cat("- P-values > 0.05 generally indicate no significant deviation from model assumptions\n")
cat("- Red areas in plots indicate potential problems\n")
```

## CHECKPOINT 4

**Model Diagnostics Summary:**

The analysis above provides several key diagnostic checks:

1. **Model Summary**: Coefficient estimates, significance tests, and model fit statistics
2. **DHARMa Residuals**: Simulation-based residual diagnostics specifically designed for GLMs
3. **Dispersion Test**: Checks if the negative binomial family appropriately handles dispersion
4. **Uniformity Test**: Tests if residuals follow expected uniform distribution
5. **Model Comparison**: Compares negative binomial vs Poisson to validate family choice

**Key Questions for Assessment:**

1. **DHARMa Diagnostics**: Do the QQ plot and residual plots look acceptable (points follow diagonal, random scatter)?
2. **Dispersion Test**: Does the dispersion test confirm appropriate modeling of variance?
3. **Model Significance**: Are the predictors (pH, Biomass, interaction) statistically significant?
4. **Model Comparison**: Was the negative binomial family justified over Poisson?

**Next Steps to Discuss:**

Based on the diagnostic results, we may need to:
- Accept the current model if diagnostics look good
- Consider model modifications if issues are detected
- Discuss biological interpretation of significant effects
- Consider model simplification if interactions are non-significant

Please review the diagnostic plots and test results, then let me know:
1. How do the DHARMa diagnostics look?
2. Are you satisfied with the model performance?
3. What would you like to focus on for interpretation or model refinement? 

### Step 2: Build model formula

**CHECKPOINT 2** Create a model with the user. 

- Ask the user to confirm their response variable. 
- Ask the user to confirm their predictor variables. Suggest that the user start with a full model with all **relevant** predictor variables. This may be a subset of all variables in the data frame. 
- Ask the user to justify the inclusion of each variable on theoretical grounds or a-priori understanding. 
- Suggest a formula for the full model and ask the user if they want to make any modifications


### Step 3: Choose a model family

Think carefully about the data type of the response variable. Consider what model family will be most suitable for the response variable. Document your reasoning for the user. 

**Default choices:** 
Continuous data positive and negative values: gaussian
Continuous values >0: gamma
Positive integers: negative binomial
Binary: binomial
Counts out of a maximum number per sample: binomial

**Proportional data:** 
Proportions or percents: Ask the user if their data can be interpreted as binomial. Often proportional data is measured as counts out of a total. If this is the case, suggest they provide the raw data and use a binomial family. 

**Special cases outside the scope of this analysis** 
If the response variable falls under the below cases, this GLM falls outside the scope of this conversation: 
True proportional data
Continuous positive values including 0
Time of day or circular response variables
Ordered or unordered categorical responses. 

**CHECKPOINT 3** 
if (choice in defaults){
    Provide suggestion to user and discuss reasoning for the choice
} else if (choice in Proportional data){
    Discuss option of modelling data with a binomial
} else if (choice in special cases){
    Inform the user that their response variable is beyond the scope of a simple glm. Suggest some follow-up reading and R packages they could investigate. 
}

### Step 4: Fit the model and check diagnostics 

Fit the model with the appropriate family. Produce diagnostic plots using the `DHARMa` package. Use these functions:

`testDispersion`
`simulateResiduals(fittedModel = m1, plot = TRUE)`

Run `vignette("DHARMa")` to provide the user more information on assessing diagnostics. 

**CHECKPOINT 4** 
Print results of diagnostics to the qmd, including figures. Render the qmd. Discuss the results with the user. Suggest next steps. 