---
title: "Linear Regression Analysis: Zooplankton Biomass vs Depth"
author: "R Workshop IMAS 2025"
date: "`r Sys.Date()`"
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: default
    code-fold: show
    fig-width: 8
    fig-height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

This document presents a linear regression analysis examining the relationship between zooplankton biomass and water depth. We'll explore how biomass changes with depth using statistical modeling and visualization techniques.

# Data Loading and Exploration

```{r load-packages}
library(tidyverse)
library(knitr)
library(kableExtra)
library(marginaleffects)
library(broom)
library(ggplot2)
```

```{r load-data}
# Load the zooplankton biomass data
dat <- read_csv("../Data/ZooBiomass.csv")

# Display basic information about the dataset
cat("Dataset dimensions:", nrow(dat), "rows,", ncol(dat), "columns\n")
head(dat) %>% 
  kable(caption = "First 6 rows of the zooplankton biomass dataset") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r data-summary}
# Summary statistics
dat %>%
  select(Depth, Biomass) %>%
  summary() %>%
  kable(caption = "Summary statistics for key variables") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

# Linear Regression Model

## Model Fitting

We fit a simple linear regression model to examine how zooplankton biomass changes with depth:

```{r fit-model}
# Fit linear model: Biomass as a function of Depth
m1 <- lm(Biomass ~ Depth, data = dat)
```

## Model Summary Table

```{r model-table}
# Create a nicely formatted summary table
model_summary <- tidy(m1, conf.int = TRUE)
model_summary %>%
  mutate(
    term = case_when(
      term == "(Intercept)" ~ "Intercept",
      term == "Depth" ~ "Depth",
      TRUE ~ term
    ),
    estimate = round(estimate, 4),
    std.error = round(std.error, 4),
    statistic = round(statistic, 2),
    p.value = case_when(
      p.value < 0.001 ~ "<0.001",
      TRUE ~ as.character(round(p.value, 3))
    ),
    conf.low = round(conf.low, 4),
    conf.high = round(conf.high, 4)
  ) %>%
  kable(
    col.names = c("Term", "Coefficient", "Std. Error", "t-statistic", "p-value", "95% CI Lower", "95% CI Upper"),
    caption = "Linear regression model summary: Biomass ~ Depth"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  add_header_above(c(" " = 5, "Confidence Interval" = 2))
```

```{r model-stats}
# Additional model statistics
model_glance <- glance(m1)
model_stats <- data.frame(
  Statistic = c("R-squared", "Adjusted R-squared", "F-statistic", "p-value (model)", "AIC", "Residual Std. Error"),
  Value = c(
    round(model_glance$r.squared, 4),
    round(model_glance$adj.r.squared, 4),
    round(model_glance$statistic, 2),
    ifelse(model_glance$p.value < 0.001, "<0.001", round(model_glance$p.value, 4)),
    round(model_glance$AIC, 2),
    round(model_glance$sigma, 4)
  )
)

model_stats %>%
  kable(caption = "Model fit statistics") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Understanding the Statistical Results (For Beginners)

**What does this table tell us?**

The summary table above shows the results of our linear regression analysis. Here's what each component means:

1. **Intercept (β₀)**: This is the predicted biomass when depth = 0 meters. In our case, it's the estimated biomass at the water surface.

2. **Depth Coefficient (β₁)**: This tells us how much biomass changes for each 1-meter increase in depth. A negative value would indicate biomass decreases with depth, while a positive value indicates it increases.

3. **Standard Error**: This measures the uncertainty in our coefficient estimates. Smaller values indicate more precise estimates.

4. **t-statistic**: This tests whether each coefficient is significantly different from zero. Larger absolute values (typically >2) suggest statistical significance.

5. **p-value**: The probability of getting these results by chance alone. Values <0.05 are typically considered statistically significant.

6. **95% Confidence Interval**: The range within which we're 95% confident the true coefficient lies.

7. **R-squared**: The proportion of variance in biomass explained by depth (ranges from 0 to 1, where 1 = perfect fit).

# Model Interpretation

```{r interpret-results}
# Extract key statistics for interpretation
intercept <- round(coef(m1)[1], 4)
slope <- round(coef(m1)[2], 4)
r_squared <- round(summary(m1)$r.squared, 4)
p_value <- round(summary(m1)$coefficients[2, 4], 4)

cat("Key findings:\n")
cat("- Intercept:", intercept, "\n")
cat("- Slope (depth effect):", slope, "\n")
cat("- R-squared:", r_squared, "\n")
cat("- P-value for depth:", ifelse(p_value < 0.001, "<0.001", p_value), "\n")
```

## Biological Interpretation

Based on our analysis:

- **Baseline biomass**: The estimated zooplankton biomass at the surface (depth = 0m) is `r round(coef(m1)[1], 3)` units.

- **Depth effect**: For every 1-meter increase in depth, zooplankton biomass `r ifelse(coef(m1)[2] > 0, "increases", "decreases")` by `r abs(round(coef(m1)[2], 4))` units on average.

- **Model fit**: Depth explains `r round(summary(m1)$r.squared * 100, 1)`% of the variation in zooplankton biomass.

- **Statistical significance**: The relationship between depth and biomass is `r ifelse(summary(m1)$coefficients[2, 4] < 0.05, "statistically significant", "not statistically significant")` (p `r ifelse(summary(m1)$coefficients[2, 4] < 0.001, "< 0.001", paste("=", round(summary(m1)$coefficients[2, 4], 3)))`).

# Predicted Values and Marginal Effects

```{r marginal-effects}
# Create predictions using marginaleffects
predictions <- predictions(m1, newdata = datagrid(Depth = seq(min(dat$Depth), max(dat$Depth), length.out = 100)))

# Plot predicted biomass against depth
ggplot(dat, aes(x = Depth, y = Biomass)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_line(data = predictions, aes(x = Depth, y = estimate), color = "red", linewidth = 1) +
  geom_ribbon(data = predictions, aes(x = Depth, y = estimate, ymin = conf.low, ymax = conf.high), 
              alpha = 0.2, fill = "red") +
  labs(
    title = "Predicted Zooplankton Biomass vs Depth",
    subtitle = "Red line shows model predictions with 95% confidence interval",
    x = "Depth (m)",
    y = "Biomass",
    caption = "Data points in blue, model predictions in red"
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )
```

# Model Verification and Diagnostic Plots

Model diagnostics help us assess whether our linear regression assumptions are met:

## Residuals vs Fitted Values

```{r residuals-fitted}
# Extract model diagnostics
model_diag <- augment(m1)

# Residuals vs Fitted plot
ggplot(model_diag, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  geom_smooth(se = FALSE, color = "blue") +
  labs(
    title = "Residuals vs Fitted Values",
    x = "Fitted Values",
    y = "Residuals"
  ) +
  theme_classic()
```

## Normal Q-Q Plot

```{r qq-plot}
# Q-Q plot for normality of residuals
ggplot(model_diag, aes(sample = .resid)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(
    title = "Normal Q-Q Plot",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  ) +
  theme_classic()
```

## Scale-Location Plot

```{r scale-location}
# Scale-Location plot
ggplot(model_diag, aes(x = .fitted, y = sqrt(abs(.resid)))) +
  geom_point(alpha = 0.6) +
  geom_smooth(se = FALSE, color = "red") +
  labs(
    title = "Scale-Location Plot",
    x = "Fitted Values",
    y = "√|Residuals|"
  ) +
  theme_classic()
```

## Cook's Distance

```{r cooks-distance}
# Cook's distance plot
ggplot(model_diag, aes(x = seq_along(.cooksd), y = .cooksd)) +
  geom_point() +
  geom_hline(yintercept = 4/nrow(dat), linetype = "dashed", color = "red") +
  labs(
    title = "Cook's Distance",
    x = "Observation Number",
    y = "Cook's Distance"
  ) +
  theme_classic()
```

## Interpretation of Diagnostic Plots

**1. Residuals vs Fitted Values:**
- **What to look for**: Points should be randomly scattered around zero with no clear patterns
- **Our results**: `r if(max(abs(model_diag$.resid)) < 2) "The residuals appear randomly scattered, suggesting the linear relationship assumption is reasonable" else "There may be some pattern in the residuals that suggests non-linearity"`

**2. Normal Q-Q Plot:**
- **What to look for**: Points should fall along the diagonal line
- **Our results**: `r if(shapiro.test(model_diag$.resid)$p.value > 0.05) "The residuals appear approximately normally distributed" else "There may be some deviation from normality in the residuals"`

**3. Scale-Location Plot:**
- **What to look for**: Points should be randomly scattered with a horizontal line
- **Our results**: This plot helps assess whether the variance of residuals is constant (homoscedasticity)

**4. Cook's Distance:**
- **What to look for**: Values above 4/n (red line) indicate potential influential observations
- **Our results**: `r if(max(model_diag$.cooksd) > 4/nrow(dat)) "Some observations may be influential" else "No highly influential observations detected"`

# Conclusions

## Summary

This analysis examined the relationship between zooplankton biomass and water depth using linear regression. The key findings are:

1. **Model Performance**: The model explains `r round(summary(m1)$r.squared * 100, 1)`% of the variation in biomass
2. **Direction of Effect**: Biomass `r ifelse(coef(m1)[2] > 0, "increases", "decreases")` with depth
3. **Statistical Significance**: The relationship is `r ifelse(summary(m1)$coefficients[2, 4] < 0.05, "statistically significant", "not statistically significant")`
4. **Model Assumptions**: The diagnostic plots suggest `r if(max(abs(model_diag$.resid)) < 2 && shapiro.test(model_diag$.resid)$p.value > 0.05) "the model assumptions are reasonably well met" else "there may be some violations of model assumptions that warrant further investigation"`

## Ecological Implications

The `r ifelse(coef(m1)[2] > 0, "positive", "negative")` relationship between depth and zooplankton biomass could be explained by several factors:

- **Light availability**: Affecting primary production and food availability
- **Temperature gradients**: Influencing metabolic rates and species distribution  
- **Oxygen levels**: Varying with depth and affecting organism survival
- **Predation pressure**: Potentially different at various depths
- **Nutrient distribution**: Affecting food web dynamics

## Recommendations for Further Analysis

1. Consider adding other environmental variables (temperature, oxygen, etc.)
2. Explore non-linear relationships if diagnostic plots suggest patterns
3. Investigate potential interactions between depth and other factors
4. Consider temporal and spatial effects in the analysis