---
title: "Backwards Stepwise Model Selection with F Ratios"
author: "R Workshop IMAS 2025"
date: "`r Sys.Date()`"
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: default
    code-fold: show
    fig-width: 8
    fig-height: 6
---

# Introduction

This document demonstrates backwards stepwise model selection using F ratios to simplify a complex model. We'll work interactively, making decisions at each checkpoint based on model diagnostics and statistical tests.

## Data and Modelling Approach

We will use zooplankton biomass data and start with a full factorial model including all main effects and interactions.

**Starting model:** `log10(Biomass) ~ Depth * Zone * Region`

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(broom)
library(car)
library(performance)
library(marginaleffects)

# Set ggplot theme
theme_set(theme_classic())
```

```{r load-data}
# Load the data
dat <- read_csv("../Data/ZooBiomass.csv")

# Examine the structure
glimpse(dat)
```

```{r data-summary}
# Summary of key variables
summary(dat[c("Region", "Zone", "Depth", "Biomass")])

# Check for any zero or negative biomass values (important for log transformation)
cat("Range of Biomass values:", range(dat$Biomass), "\n")
cat("Any zero or negative values?", any(dat$Biomass <= 0), "\n")
```

# Step 1: Fit the Full Model and Check Diagnostics

```{r full-model}
# Fit the full model with 3-way interaction
full_model <- lm(log10(Biomass) ~ Depth * Zone * Region, data = dat)

# Model summary
summary(full_model)
```

```{r anova-table}
# ANOVA table
anova(full_model)
```

## Model Diagnostics

```{r diagnostic-plots, fig.height=8}
# Create diagnostic plots
par(mfrow = c(2, 2))
plot(full_model)
par(mfrow = c(1, 1))
```

```{r additional-diagnostics}
# Additional diagnostic tests
cat("Shapiro-Wilk test for normality of residuals:\n")
shapiro.test(residuals(full_model))

cat("\nBreusch-Pagan test for homoscedasticity:\n")
bptest <- car::ncvTest(full_model)
print(bptest)

# Residuals vs fitted with smoother
library(ggplot2)
diagnostic_df <- data.frame(
  fitted = fitted(full_model),
  residuals = residuals(full_model),
  standardized_residuals = rstandard(full_model)
)

ggplot(diagnostic_df, aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.6) +
  geom_smooth(se = FALSE, color = "red") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Residuals vs Fitted Values",
    x = "Fitted Values",
    y = "Residuals"
  ) +
  theme_classic()
```

```{r qq-plot}
# Q-Q plot with ggplot
ggplot(diagnostic_df, aes(sample = standardized_residuals)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(
    title = "Q-Q Plot of Standardized Residuals",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  ) +
  theme_classic()
```

```{r model-performance}
# Model performance metrics
check_model(full_model, check = c("normality", "homogeneity"))
```

## **CHECKPOINT 1**: Model Diagnostics Assessment

Based on the diagnostic plots and tests above, we need to evaluate:

1. **Normality**: Do the residuals follow a normal distribution?
2. **Homoscedasticity**: Is the variance constant across fitted values?
3. **Model Structure**: Is the log10 transformation appropriate, or should we consider a GLM approach?

### Questions for Decision:
- Are you satisfied with the current model structure and log10 transformation?
- Do we need to consider alternative approaches (e.g., GLM with Gamma family)?
- Should we proceed with this model structure for stepwise selection?

**Please review the diagnostics above and confirm whether to proceed with the current model structure or make adjustments.**

---

# Step 2: Test the 3-Way Interaction

Now we begin backwards stepwise model selection by testing whether we can remove the highest-order term (the 3-way interaction: `Depth:Zone:Region`).

```{r step2-models}
# Current full model (with 3-way interaction)
full_model <- lm(log10(Biomass) ~ Depth * Zone * Region, data = dat)

# Reduced model (without 3-way interaction)
reduced_model <- lm(log10(Biomass) ~ Depth + Zone + Region + 
                    Depth:Zone + Depth:Region + Zone:Region, data = dat)

# Alternative way to fit the same reduced model
# reduced_model <- lm(log10(Biomass) ~ (Depth + Zone + Region)^2, data = dat)
```

```{r step2-comparison}
# Compare the models using F-test
f_test <- anova(reduced_model, full_model)
print(f_test)

cat("\n=== Model Comparison Summary ===\n")
cat("Full model AIC:", AIC(full_model), "\n")
cat("Reduced model AIC:", AIC(reduced_model), "\n")
cat("AIC difference (reduced - full):", AIC(reduced_model) - AIC(full_model), "\n")
cat("\nF-test p-value for 3-way interaction:", f_test$`Pr(>F)`[2], "\n")
```

```{r step2-effect-sizes}
# Look at the 3-way interaction term in the full model
cat("=== 3-Way Interaction Coefficient ===\n")
full_summary <- summary(full_model)
interaction_coef <- full_summary$coefficients["Depth:ZoneOffshore:RegionReference", ]
print(interaction_coef)

cat("\n=== Effect Size Context ===\n")
cat("3-way interaction coefficient:", interaction_coef[1], "\n")
cat("Standard error:", interaction_coef[2], "\n")
cat("t-value:", interaction_coef[3], "\n")
cat("p-value:", interaction_coef[4], "\n")
```

```{r step2-anova-comparison}
# Show ANOVA tables for both models
cat("=== Full Model ANOVA ===\n")
print(anova(full_model))

cat("\n=== Reduced Model ANOVA ===\n")
print(anova(reduced_model))
```

```{r step2-r-squared}
# Compare R-squared values
full_r2 <- summary(full_model)$r.squared
reduced_r2 <- summary(reduced_model)$r.squared
adj_full_r2 <- summary(full_model)$adj.r.squared
adj_reduced_r2 <- summary(reduced_model)$adj.r.squared

cat("=== Model Fit Comparison ===\n")
cat("Full model R²:", round(full_r2, 4), "\n")
cat("Reduced model R²:", round(reduced_r2, 4), "\n")
cat("R² difference:", round(full_r2 - reduced_r2, 4), "\n")
cat("\nFull model Adjusted R²:", round(adj_full_r2, 4), "\n")
cat("Reduced model Adjusted R²:", round(adj_reduced_r2, 4), "\n")
cat("Adjusted R² difference:", round(adj_full_r2 - adj_reduced_r2, 4), "\n")
```

## **CHECKPOINT 2**: Decision on 3-Way Interaction

Based on the analysis above, we need to decide whether to remove the 3-way interaction (`Depth:Zone:Region`).

### Decision Criteria:
1. **F-test p-value**: Is p > 0.05? (suggests we can remove the term)
2. **AIC comparison**: Does the reduced model have lower AIC? (suggests better model)
3. **Effect size**: Is the coefficient practically meaningful?
4. **R² change**: Is the reduction in explained variance acceptable?

### Interpretation Guide:
- **If F-test p > 0.05**: The 3-way interaction is not statistically significant → **Remove it**
- **If F-test p ≤ 0.05**: The 3-way interaction is statistically significant → **Keep it**
- **AIC difference**: Lower AIC indicates better model (accounting for complexity)
- **R² change**: Small change suggests the interaction doesn't explain much variance

**Please review the results above and decide: Should we remove or keep the 3-way interaction?**

